Simulating High-Dimensional Multivariate Data

### 2020-10-24 010-background.Rmd
Revising background

# Background and notation

Formally, this is our target:

We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data. The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's $\tau$ or Spearman's $\rho$, and approximately Pearson product-moment correlation.More formally, our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
The algorithm must scale to a large number of simulation replicates,$N$ for high-dimensional multivariate constructions (dimension $d$ in the tens or hundred thousands.

Ultra High Dimensional multivariate modeling and simulation desired properties, both basic properties (BP) and scaleability properties (SP). 
With our application in mind and seeking a fairly general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from @Nik13a):

* BP1: Wide range of dependence, allowing both positive and negative dependence
* BP2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

Moreover, the simulation method must scale to high dimensions:

@@

@@

Simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes [@BE07; @Schissler2018]. 
RNA-sequencing analysis has garnered much interest in the statistical literature (refs). 
Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). 
An alternative strategy is to think *generatively* about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes.
Our methodology provides another toolkit to generate data by considering how the data arise in the experimental setting.

@@

@@

We present an general-purpose, scalable multivariate simulation algorithm. The crux of the method is construction of a Gaussian copula. This idea is well known [refs]. We chose a Gaussian copula among the many available copula functions primarily for two reasons: 1) the computationally convient closed expression to convert among the most common dependency measures when the margins are Gaussian and 2) the availability of high-performance multivariate normal simulators in R ('mvnfast' ref).

@@

## Matching Pearson correlations

In the NORTA, it is well known that blah

## Discrete marginal considerations

<!-- 
For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.

\noindent Kendall's $\tau$ is the probability of concordant pairs minus the probability of discordant pairs and is given compactly as

\begin{equation}
  (\#eq:tau)
  \tau(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
\end{equation}

-->

## Marginal-dependent bivariate correlation bounds

Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
Yet not all multivariate simulation approaches obtain these bounds [for example REF].
Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].

>>>>

# Background and notation {#background}

The article presents several algorithms to work with high-dimesional multivariate data, but all `bigsimr` algorithms were originally designed to support a single task:
to generate random vectors drawn from given marginal distributions and component-wise dependency metrics.
Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, where $d$ can be very large (possibly millions of variables).

When designing this methodology, we developed the following properties to guide our effort.
We divide the properties into two categories: (1) basic properties (**BP**) and "scaleability" properties (**SP**). 
The BPs are adapted from an existing criteria due to @Nik13a.
Our simulation strategy should allow:

* BP1: Wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

Moreover, the simulation method must **scale** to high dimensions:

@@

To fix ideas and provide example applications enabled via `bigsimr`, we the next section describes our motivating data that originally inspired the authors` interest in creating this methodology.

@@

Simulating high-dimensional non-normal data motivates this work --- in pursuit of modeling RNA-sequencing data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Often researchers posit a negative-binomial (NB) model as counts are often over-dispersed that a Poisson model would suggest.
Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].

@@

To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:

\begin{equation}
(\#eq:convertKendall)
r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's $\rho$ [@K58],

\begin{equation}
(\#eq:convertSpearman)
\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
\end{equation}

*Marginal-dependent bivariate correlation bounds*. 
Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation.
This is a consequence of the bounds of bivariate distribution functions --- the *Frechet-Hoeffding bounds* [@Nelsen2007]). 
Denote the pointwise upper bound as $M(y_i, y_{i^\prime})$ These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. 
Yet not all multivariate simulation approaches obtain these bounds [for example REF].
Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration [@Li2019gpu].

For discrete marginals, however, one could perform a rescaled version of $\rho_s$ for random variables $X,Y$ with pmfs (or pdfs) $p(x)$ and $q(y)$, respectively.

@@

The crux of the method is construction of a Gaussian copula. This idea is well known [refs]. We chose a Gaussian copula among the many available copula functions primarily for two reasons: 1) the computationally convient closed expression to convert among the most common dependency measures when the margins are Gaussian and 2) the availability of high-performance multivariate normal simulators in R ('mvnfast' ref).

@@

### 2020-10-24 020-simulation-algorithm.Rmd
revise introduction AGS
Revising background

# Algorithms

```{r previewChapter020, echo = FALSE, eval = FALSE}
bookdown::preview_chapter('020-simulation-algorithm.Rmd')
```

@@

@@

We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs]. 
To do this, we take advantage of a closed form relationship [ref?] between Kendall's $\tau$ and Pearson's correlation coefficient for bivariate normal random variables:

\begin{equation}
(\#eq:convertKendall)
r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's $\rho$ [@K58],

\begin{equation}
(\#eq:convertSpearman)
\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
\end{equation}

@@

## Random vector generation `bigsimr::rvec` algorithm

>>>>

# Algorithms {#algorithms}

```{r ch020-preview, echo = FALSE, eval = FALSE}
bookdown::preview_chapter('020-simulation-algorithm.Rmd')
```

@@

`mvnfast` High-performance multivariate normal simulator [@Fasiolo2016]

`nortaRA` Implements exact Pearson matching (step 2 in NORTA) [@Chen2001]

@@

We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved [refs].

@@

## Random vector generation `bigsimr::rvec` algorithm {#rand-vec-gen}

### 2020-10-24 040-monte-carlo-evaluation.Rmd
Revising background

# Monte Carlo evaluations

@@

## Simulation I: Bivariate simulation using rank-based correlation

### Bivariate Normal

* Let's simulate a bivariate normal and check our correlation matching
  performance as N increases.
* Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ )
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

@@

## Bivariate Gamma

* Similarly, let's check the performance for a non-symmetric continuous distribution: a
  standard (rate =1) bivariate gamma.
* Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

@@

## Bivariate Negative Binomial

* Let's check the performance for a discrete distribution: a bivariate negative binomial
* Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ )
* We vary $\rho$ across the entire possible range of correlations for each
  correlation type.

>>>>

# Monte Carlo evaluations {#simulations}

@@

## Bivariate experiments

*Bivariate Normal*.
Let's simulate a bivariate normal and check our correlation matching performance as N increases.
Here we have BVN( $\mu_1 = \mu_2 = 10, \rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.

@@

*Bivariate Gamma*. 
Similarly, let's check the performance for a non-symmetric continuous distribution: a standard (rate =1) bivariate gamma.
Here we have a Bivariate Gamma with $shape_1 = shape_2 = 10, \rho_{type}$.
We vary $\rho$ across the entire possible range of correlations for each correlation type.

@@

*Bivariate Negative Binomial*. 
Let's check the performance for a discrete distribution: a bivariate negative binomial.
Here we have Bivariate Negative Binomial ( $prob_1 = prob_2 = 0.5, size_1 = size_2 = 4,\rho_{type}$ ).
We vary $\rho$ across the entire possible range of correlations for each correlation type.
